---
title: "Construct Phenoclimatic Models"
author: "Natalie Love"
date: "1/22/2020"
output: html_document
---
##Load Packages
```{r}
library(tidyverse)
library(broom) #for tidying model output
library(car) #if you want to use type 3 sums of squares
library(leaflet)
library(visreg) #for visualizing partial effect plots
```

##Load data
Load the dataframe that contains the climate data associated with each record
```{r}
esca_pheno_climate <- read_csv("esca_pheno_climate.csv", col_types = cols(
  id = col_double(),
  institutionCode = col_character(),
  collectionCode = col_character(),
  ownerInstitutionCode = col_character(),
  basisOfRecord = col_character(),
  occurrenceID = col_character(),
  catalogNumber = col_character(),
  otherCatalogNumbers = col_character(),
  kingdom = col_character(),
  phylum = col_character(),
  class = col_character(),
  order = col_character(),
  family = col_character(),
  scientificName = col_character(),
  taxonID = col_double(),
  scientificNameAuthorship = col_character(),
  genus = col_character(),
  specificEpithet = col_character(),
  taxonRank = col_character(),
  infraspecificEpithet = col_character(),
  recordedBy = col_character(),
  associatedCollectors = col_character(),
  eventDate = col_character(),
  year = col_double(),
  month = col_double(),
  day = col_double(),
  startDayOfYear = col_double(),
  endDayOfYear = col_double(),
  verbatimEventDate = col_character(),
  occurrenceRemarks = col_character(),
  habitat = col_character(),
  substrate = col_character(),
  verbatimAttributes = col_character(),
  fieldNumber = col_character(),
  informationWithheld = col_character(),
  dataGeneralizations = col_character(),
  dynamicProperties = col_character(),
  associatedTaxa = col_character(),
  reproductiveCondition = col_character(),
  establishmentMeans = col_character(),
  cultivationStatus = col_character(),
  lifeStage = col_character(),
  sex = col_character(),
  individualCount = col_double(),
  preparations = col_character(),
  country = col_character(),
  stateProvince = col_character(),
  county = col_character(),
  municipality = col_character(),
  locality = col_character(),
  locationRemarks = col_character(),
  localitySecurity = col_double(),
  localitySecurityReason = col_character(),
  decimalLatitude = col_double(),
  decimalLongitude = col_double(),
  geodeticDatum = col_character(),
  coordinateUncertaintyInMeters = col_double(),
  verbatimCoordinates = col_character(),
  georeferencedBy = col_character(),
  georeferenceProtocol = col_character(),
  georeferenceSources = col_character(),
  georeferenceVerificationStatus = col_character(),
  georeferenceRemarks = col_character(),
  minimumElevationInMeters = col_double(),
  maximumElevationInMeters = col_double(),
  minimumDepthInMeters = col_double(),
  maximumDepthInMeters = col_double(),
  verbatimDepth = col_character(),
  verbatimElevation = col_character(),
  disposition = col_character(),
  language = col_character(),
  recordEnteredBy = col_character(),
  modified = col_character(),
  sourcePrimaryKey_dbpk = col_character(),
  collId = col_double(),
  recordId = col_character(),
  references = col_character()
))
View(esca_pheno_climate)
```



2. Simple plots
These can be used to detect problems with the data. For example there seems to be one outlier temperature value
```{r}
#Using base graphics
plot(esca_pheno_climate$startDayOfYear ~ esca_pheno_climate$springTmean_norm)
plot(esca_pheno_climate$startDayOfYear ~ esca_pheno_climate$winterPPT_norm)

#Using ggplot2
ggplot(data = esca_pheno_climate, aes(x = springTmean_norm, y = startDayOfYear)) + geom_point() + theme_bw()
ggplot(data = esca_pheno_climate, aes(x = winterPPT_norm, y = startDayOfYear)) + geom_point() + theme_bw()
```

3. Look at a map
Leaflet is just one way to make a map, but it is interactive rather than static which is nice
```{r}
leaflet(esca_pheno_climate) %>%
  addTiles() %>%
  addMarkers(lng = esca_pheno_climate$decimalLongitude, lat = esca_pheno_climate$decimalLatitude)
```

##Construct Phenoclimatic Models
```{r}
#simple regressions
#spring temp - 30-year normal
esca_temp_model <- lm(startDayOfYear ~ springTmean_norm, data = esca_pheno_climate)
summary(esca_temp_model)

par(mfrow = c(2, 2))
plot(esca_temp_model)

#winter PPT - 30-year normal
esca_ppt_model <- lm(startDayOfYear ~ winterPPT_norm, data = esca_pheno_climate)
summary(esca_ppt_model)

#multiple regression and interaction
esca_climate_model <- lm(startDayOfYear ~ springTmean_norm + winterPPT_norm, data = esca_pheno_climate)
summary(esca_climate_model)

esca_climate_inter <- lm(startDayOfYear ~ springTmean_norm + winterPPT_norm + springTmean_norm*winterPPT_norm, data = esca_pheno_climate)
summary(esca_climate_inter)
```

##Visualize model results
There are multiple ways to visualize model output. The package visreg makes it pretty easy to visualize partial effect plots from model objects. This is especially useful for multivariate regressions and interactions. ggplot can be used to visualize simple regressions.
```{r}
visreg(esca_climate_model, xvar = "winterPPT_norm", xlab = "Cumulative winter PPT - 30yr normal", ylab = "DOY")
visreg(esca_climate_model, xvar = "springTmean_norm", xlab = "Spring Tave - 30yr Normal", ylab = "DOY")

#for interactions
visreg(esca_climate_inter, xvar = "winterPPT_norm", by = "springTmean_norm", overlay = TRUE, xlab = "Cumulative winter PPT - 30yr normal", ylab = "DOY")
```

##Export model results
Some simple code can be used to export model summaries as csv files
```{r}
#example using the temperature model
write_csv(tidy(esca_temp_model), "esca_temp_model.csv")
```

